Cyclone: Replication Middleware for NVM Clusters

### 2016-09-08 paper/paper.tex
Additions to paper.

\subsection{Network Stack}
Replication protocols are expensive to operate - fact that has been underlined in
multiple pieces of research [FARM, in a box]. It would therefore appear that
replication would become the bottleneck in Cyclone. We observed however that
previous work did not implement replication around a high performance network
stack possibly explaining why they chose both high performance hardware (40
GigE+ NICS, custom FPGA designs) as well as specific high performance but
restrictive software APIs (one sided RDMA with metadata co-location, application
compiled down to the FPGA). As a counterpoint, we designed Cyclone around DPDK
[dpdk] a high performance user-space network stack. DPDK provides convenient
access to packets arriving at an ethernet NIC by moving them into packet buffers
with the contained cache lines also directly placed into the last level cache of
the CPU during DMA.

Cyclone separates the available cores on the machine into cores dedicated to
work related to the consensus protocol and cores decidated to running
application code. The division is configurable but Fig [] shows the division for
our testbed. Each RAFT core runs an \emph{independent} instance of the RAFT
protocol (Section~\ref{sec:ind}). RAFT cores pass work to the application cores
through lock-free double ended queues. A RAFT core dispatches an incoming RPC
call for execution simultaneously with sending it out for replication. As
explained in Section~\ref{sec:decouple} this is beneficial for reducing latency
to completion.

Each RAFT core has a dedicated NIC queue pair. The input queue receives RPC
calls from clients as well as RAFT protocol related traffic. The output queue
only carries RAFT protocol related traffic. Each application core has a
dedicated \emph{output} queue on the NIC. When completed, the response to the
RPC call is sent from the application core via its dedicated output queue. This
division of NIC queues eliminates any need for synchronization between RAFT and
application cores when accessing the NIC.

\subsection{Zero Copy Replication}
Consensus protcols such as RAFT usually have a simple common case path. Once a
quorum leader is elected, the leader's primary job is to transmit new entries
appended to the log to its followers. The primary job of the followers is to
receive these entries and send acknowledgments to the leader. An entry is
considered committed when the leader receives an acknowledgment for an entry
from a majority of the quorum. Followers receive intimation of committed entries
as piggybacked information on subsequent replication messages.

Our aim when designing Cyclone was to be relatively insensitive to the size of
the RPC call - i.e. the size of the RAFT log entry being replicated. Since
Cyclone is targeted towards general purpose applications we intended to support
a range of RPC call sizes without a degradation in performance caused by copying
the RPC call contents for transmission to each follower replica. Hence, we built
a zero copy mechanism for common case of a leader replica sending the same log
entry to a set of follower replicas. To achieve our zero copy goal we explot the
fact that DPDKs packet buffers allow prepending a data without physically moving
the data in the packet buffer. Fig [] shows how the layout of a packet buffer
changes at the leader during replication. The received packets contains only the
RPC call from the client. We prepend a RAFT related header (containing the
previous log index, previous log term, leader's term and leader's commit
point) to packet buffer itself. To this RAFT header we then prepend an IPv4
header. Although we use the raw packet interface from DPDK and do not use IP
based routing, we set IP addresses in a dummy header to use 5-tuple steering in
order to place the packet in the right queue at the receiving NIC. This
composite packet buffer is then sent to every replica. In order to properly
route the packet through switches, we allocate a small buffer containing an
ethernet header for each follower replica and \emph{chain} it in front of the
prepared packet - ensuring we do not need to change the contents for every
replica.

\subsection{Batching}
Although the common case path in RAFT is straightforward, it still adds many
hundreds of cycles for each log entry to be processed. This adds up to a few
microseconds, a significant amount in the face of sub 10 us latency between
nodes with DPDK. We amortize this overhead when operating under load by batching
multiple RPC calls together and treating the batch as a single RAFT log entry.
DPDK by itself can efficiently batch incoming packets under load - allowing the
user to easily receive a burst (currently limited to 32) packets rather than a
single one for a polling call. DPDK further allows chaining packet buffers
together and treating the chain as a single packet. This is convenient for our
zero copy approach and we chain packet mbufs together to form a single packet
as shown in Fig []. Since the entire chain is sent as a single packet when being
replicated, the length of a chain is limited by the size of the coalesced
packet, that must fit into an ethernet jumbo frame of approximately 9K bytes.

@@

\section{Putting it together - A Durable Replicated Hash Table}
\label{sec:example}
We now demonstrate how Cyclone can be used in practise by putting together an
example of a linearizable hash table - a useful and common data structure
[memcached]. The baseline is a durable hash table written in NVML that runs on a
single machine in client-server mode. The server side maintains the durable
open-addressed hash table and services client RPC calls to insert and delete
key-value pairs and lookup keys. Such a hash table is simple to derive from a
volatile hash table and readily available in NVML [nvml.io] as an example [src
  code link]. To use that (non client-server) example with Cyclone we add a
simple RPC call format, manually marshall and demarshall the (at most two)
arguments to the RPC call in an RPC call handler we wrote (a few tens of lines
of code) and wrote a simple client driver program (a few tens of additional
lines of code).

In order to use the hash table with Cyclone, we simply need to to declare the RPC
call handler as the point of entry to Cyclone at the server end - shown in
Figure []. The client end issues RPC calls via Cyclone's client side library as
shown in Figure []. Crucially neither side involves fault awareness or fault
recovery code - a key goal for Cyclone.

@@

>>>>

\subsection{Network Stack}
Replication protocols are expensive to operate - fact that has been underlined
in multiple pieces of research [FARM, in a box]. It would therefore appear that
replication would become the bottleneck in Cyclone. We observed however that
previous work did not implement replication around a high performance network
stack possibly explaining why they chose both high performance hardware (40
GigE+ NICS, custom FPGA designs) as well as specific software APIs (one sided
RDMA with metadata co-location, application compiled down to the FPGA) that
bypass the network stack. As a counterpoint, we designed Cyclone around DPDK
[dpdk] a high performance user-space network stack. DPDK provides convenient
access to packets arriving at an ethernet NIC by moving them into packet buffers
with the contained cache lines also directly placed into the last level cache of
the CPU during DMA.

Cyclone separates the available cores on a machine into cores dedicated to work
related to the consensus protocol and cores decidated to running application
code. The division is configurable. Each RAFT core runs an \emph{independent}
instance of the RAFT protocol (Section~\ref{sec:ind}). RAFT cores pass work to
the application cores through lock-free double ended queues. A RAFT core
dispatches an incoming RPC call for execution simultaneously with sending it out
for replication. As explained in Section~\ref{sec:decouple} this is beneficial
for reducing latency to completion.

Each RAFT core has a dedicated NIC queue pair. The input queue receives RPC
calls from clients as well as RAFT protocol related traffic. The output queue
only carries RAFT protocol related traffic. Each application core has a
dedicated \emph{output} queue on the NIC. When completed, the response to the
RPC call is sent from the application core via its dedicated output queue. This
division of NIC queues eliminates any need for synchronization between RAFT
cores and application cores when accessing the NIC.

In the event that the machine has multiple NICs Cyclone binds each RAFT or
application core to a specific NIC and picks the necessary queues from that
NIC. Spreading the network demands from the cores across the NICs in this manner
allows Cyclone to scale with multiple NICs when available.

\subsection{Zero Copy Replication}
Consensus protcols such as RAFT usually have a simple common case path. Once a
quorum leader is elected, the leader's primary job is to transmit new entries
appended to the log to its followers. The primary job of a follower is to
receive these entries, append them to its log and send acknowledgments to the
leader. An entry is considered committed when the leader receives an
acknowledgment for an entry from a majority of the quorum. Followers receive
intimation of committed entries as piggybacked information on subsequent
replication messages.

Our aim when designing Cyclone was to be relatively insensitive to the size of
the RPC call - i.e. the size of the RAFT log entry being replicated. Since
Cyclone is targeted towards general purpose applications we intended to support
a range of RPC call sizes without a degradation in performance caused by copying
the RPC call contents for transmission to each follower replica. Hence, we built
a zero copy mechanism for the common case of a leader replica sending the same
log entry to a set of follower replicas. To achieve our zero copy goal, we
exploit the fact that DPDKs packet buffers allow prepending data without
physically moving the data in the packet buffer. Fig [figures/packets.bmp] shows
how the layout of a packet buffer changes at the leader during replication. The
received packets contains only the RPC call from the client. We prepend a RAFT
related header (containing the previous log index, previous log term, leader's
term and leader's commit point) to packet buffer itself. To this RAFT header we
then prepend an IPv4 header. Although we use the raw packet interface from DPDK
and do not use IP based routing, we set IP addresses in a dummy header to use
5-tuple steering in order to place the packet in the right queue at the
receiving NIC. This composite packet buffer is then sent to every replica. In
order to properly route the packet through switches, we allocate a small buffer
containing an ethernet header for each follower replica and \emph{chain} it in
front of the prepared packet - ensuring we do not need to change the contents
for every replica.

\subsection{Batching}
Although the common case path in RAFT is straightforward, it still adds many
hundreds of cycles for each log entry to be processed. This adds up to a few
microseconds, a significant amount in the face of sub 10 us network latency
between nodes with DPDK. We amortize this overhead when operating under load by
batching multiple RPC calls together and treating the batch as a single RAFT log
entry.  DPDK by itself can efficiently batch incoming packets under load -
allowing the user to easily receive a burst (currently limited to 32) of packets
rather than a single one for a polling call. DPDK further allows chaining packet
buffers together and treating the chain as a single packet. This is convenient
for our zero copy approach and we chain packet mbufs together to form a single
packet as shown in Fig [figures/chain.bmp]. Since the entire chain is sent as a
single packet when being replicated, the length of a chain is limited by the
size of the coalesced packet, that must fit into an ethernet jumbo frame of
approximately 9K bytes.

@@

\section{Putting it together - A Durable Replicated Hash Table}
\label{sec:example}
We now demonstrate how Cyclone can be used in practice by putting together an
example of a linearizable hash table - a useful and common data structure
[memcached]. The baseline is a durable hash table written in NVML that runs on a
single machine in client-server mode. The server side maintains the durable
open-addressed hash table and services client RPC calls to insert and delete
key-value pairs and lookup keys. Such a hash table is simple to derive from a
volatile hash table and readily available in NVML [nvml.io] as an example [src
  code link]. To use that (non client-server) example with Cyclone we add a
simple RPC call format, manually marshall and demarshall the (at most two)
arguments to the RPC call in an RPC call handler we wrote (a few tens of lines
of code) and wrote a simple client driver program (a few tens of additional
lines of code).

In order to use the hash table with Cyclone, we simply need to to declare the
RPC call handler as the point of entry to Cyclone at the server end - shown in
Figure~\ref{fig:server_api}. The client end issues RPC calls via Cyclone's
client side library as shown in Figure~\ref{fig:client_api}. Crucially neither
side involves fault awareness or fault recovery code - a key goal for Cyclone.

@@

\begin{figure}
{ \scriptsize
\begin{verbatim}
int exec_rpc_call(void * input, int input_size, void *output)
{ 
  ...
}

struct rpc_callbacks_t callbacks = {
.exec_call = exec_rpc_call,
...
};

main()
{
  ...
  cyclone_start(callbacks, ...);
  ...
}
\end{verbatim}
}
\caption{Server side API}
\label{fig:server_api}
\end{figure}

\begin{figure}
{ \scriptsize
\begin{verbatim}
int make_rpc_call(..., void *data, int size, int rpc_number);
int get_last_rpc_number(...);
main()
{
  ...
  int rpc_number = get_last_rpc_number(...);
  while(true) {
    ...
    make_rpc_call(..., data, size, rpc_number++);

  }
  ...
}
\end{verbatim}
}
\caption{Client side API}
\label{fig:client_api}
\end{figure}
